---
title: "Data Processing"
author: "Team 2"
date: "2025-11-03"
output: pdf_document
---


## Data reading

```{r}

data <- read.csv("00_ProstateCancer_Data.csv", header=T)
head(data)
str(data)
summary(data)

```



## Dataset Variables


### Initial Diagnosis


**Hasta_ID (Categorical)**: Patient ID


**Yas (Discrete)**: Age


**Tani_Tarihi (Date)**:	Diagnosis Date


**PSA_Tani (Continuous)**: Serum Prostate-Specific Antigen (PSA) level at diagnosis (ng/mL)


**Klinik_Evre	(Ordinal/Categorical)**: Clinical cT-Stage determined by pre-treatment examinations (cT1c < cT2a < cT2b < cT2c < cT3a < cT3b for increasing extent of tumor invasion)


**Biyopsi_Gleason	(Ordinal/Categorical)**: Biopsy Gleason Score (3+3 < 3+4 < 4+3 < 3+5 < 4+4 < 4+5 < 5+4 < 5+5, higher score indicates higher aggressiveness)


**Risk_Grubu (Ordinal/Categorical)**: Risk Group Classification (1 for Low, 2 for Intermediate, 3 forHigh)	


### Risk Factors


**Albumin	(Continuous)**: ASerum albumin level (g/dL). Indicator of nutritional status and systemic health


**Lenfosit (Discrete)**: Lymphocyte (Immune system component) Count


**CRP (Continuous)**: C-Reactive Protein (mg/L). Indicator of inflammation


**NLR	(Continuous)**: Neutrophil-to-Lymphocyte Ratio. A prognostic indicator for systemic inflammation and cancer aggressiveness.


**CALLY_Index	(Continuous)**: CALLY Index. A composite index, likely related to inflammation or blood components.


**Komorbidite_Skor (Ordinal/Categorical)**: Comorbidity Score indicating the severity of other co-existing chronic diseases ( 0 (No comorbidities) < ... < 5 (Severe comorbidities))


### Treatment Information


**Tedavi_Tipi	(Categorical)**: Main Treatment Type received (1 for Radical Prostatectomy, 2 for Radiotherapy/RT, 3 for Hormone Therapy, 4 for Combination of Radiotherapy and Hormone Therapy)


**Tedavi_Tarihi	(Date)**: Treatment Date


**RT_Dozu (Continuous)**: Total Radiation Dose (in Gy), if radiotherapy was performed


**ADT_Tipi (Categorical)**: Androgen Deprivation Therapy (ADT, hormone therapy) Type used


**ADT_Suresi (Continuous)**: ADT(hormone therapy) Duration


### Pathological Markers


**Patolojik_Evre (Ordinal/Categorical)**: Final Tumor Pathological Stage determined after surgery on the removed tissue (pT2a < pT2b < pT2c < pT3a < pT3b < pT4, NaN indicates patient did not undergo surgery)


**Cerrahi_Sinir	(Binary/Categorical)**: Surgical Margin Status indicating if cancer cells were present at the edge of the removed tissue. Crucial for recurrence prediction (0: Negative, 1: Positive, NaN indicates patient did not undergo surgery). 


**Final_Gleason	(Ordinal/Categorical)**: Final Gleason Score confirmed from the final excised tissue (3+3 < 3+4 < 4+3 < 3+5 < 4+4 < 4+5 < 5+4 < 5+5, higher score indicates higher aggressiveness)


### Follow-up & Outcomes


**PSA_Nadir	(Continuous)**: The lowest PSA level reached after treatment (ng/mL). A lower nadir generally indicates better treatment success


**PSA_Takip_3ay / 6ay / 12ay (Continuous)**: Follow-up PSA levels (ng/mL) measured at at 3/6/12 Months


**BCR_Durum (Binary/Categorical)**: Biochemical Recurrence (BCR) Status	whether the PSA level rise above a recurrence threshold? (True for Recurrence occurred, False for no Recurrence occurred)


**BCR_Tarihi (Date)**: Date when biochemical recurrence was confirmed


**Metastaz_Durum (Binary/Categorical)**: Metastasis Status whether distant metastasis occur during follow-up? (0 for No, 1 for Yes)


**Metastaz_Tarihi	(Date)**: Date when metastasis was confirmed


**Son_Durum	(Binary/Categorical)**: Patient's Survival Status at the last follow-up (0 for Alive, 1 for Deceased)


**Son_Takip_Tarihi (Date)**: Date of the last recorded patient information.



## Data handling 1: Remove variables that we will not use for analysis

```{r}

cols1 <- c("Tani_Tarihi", "Tedavi_Tarihi", "PSA_Takip_3ay", "PSA_Takip_6ay", "PSA_Takip_12ay", "BCR_Tarihi", "Metastaz_Tarihi", "Son_Takip_Tarihi")

data <- data[, !(names(data) %in% cols1)]

```



## Data handling 2: Turning all categorical variables to factors

```{r}

cols2 <- c("Klinik_Evre", "Biyopsi_Gleason", "Risk_Grubu", "Komorbidite_Skor", "Tedavi_Tipi", "ADT_Tipi", "Patolojik_Evre", "Cerrahi_Sinir", "Final_Gleason", "BCR_Durum", "Metastaz_Durum", "Son_Durum")

data[cols2] <- lapply(data[cols2], as.factor)

```



## Data handling 3: Changing variable names

```{r}

names(data) <- c("Patient ID", "Age", "PSA_before", "CTstage", "GleasonScore_before", "RiskClass", "Albumin", "Lymphocyte", "CRP", "NLR", "CallyIndex", "ComorbidityScore", "Treatment", "RadiationDose", "HormoneType", "HormonDuration" ,"TumorSize", "MarginStatus", "GleasonScore_after", "PSA_after", "BCR", "Metastasis", "Survival")

```

```{r}

str(data)

```



## Data Visualization 1: Frequency of Treatment Type (Treatment)

```{r}

sum(is.na(data$Treatment))

table(data$Treatment)
round(prop.table(table(data$Treatment)), 3)

frequencies <- table(data$Treatment)
barplot(
    frequencies,
    main = "Frequency of Treatment Type (Treatment)",
    xlab = "Treatment Type",
    ylab = "Number of Patients",
    col = rainbow(4),
    ylim = c(0, max(frequencies) * 1.1)
    )

```



## Data Visualization 2: Frequency of Biochemical Recurrence Status (BCR)

```{r}

sum(is.na(data$BCR))

table(data$BCR)
round(prop.table(table(data$BCR)), 3)

frequencies <- table(data$BCR)
barplot(
    frequencies,
    main = "Frequency of Biochemical Recurrence Status",
    xlab = "Recurrence Status",
    ylab = "Number of Patients",
    col = c("blue", "red"),
    ylim = c(0, max(frequencies) * 1.1)
    )

```



## Data Visualization 3: Frequency of Survival Status (Survival)

```{r}

sum(is.na(data$Survival))

table(data$Survival)
round(prop.table(table(data$Survival)), 3)

frequencies <- table(data$Survival)
barplot(
    frequencies,
    main = "Frequency of Survival Status",
    xlab = "Survival Status",
    ylab = "Number of Patients",
    col = c("blue", "red"),
    ylim = c(0, max(frequencies) * 1.1)
    )

```



## Data Visualization 4: Serum Prostate-Specific Antigen (PSA) level at diagnosis & after treatment

```{r}

PSA <- c("PSA_before", "PSA_after")
sapply(data[, PSA], function(x) sum(is.na(x)))

summary(data[, PSA])

par(mfrow = c(1, 2))

boxplot(data[["PSA_before"]],
          main = "Box Plot of PSA level at diagnosis",
          col = "grey")

boxplot(data[["PSA_after"]],
          main = "Box Plot of PSA level after treatment",
          col = "green")

```


The PSA level significantly decreased after treatment.


## Data Visualization 5: Summary & Boxplots, Barplot of Risk Factors

```{r}

risk_factors <- c("Albumin", "Lymphocyte", "CRP", "NLR", "CallyIndex", "ComorbidityScore")
sapply(data[, risk_factors], function(x) sum(is.na(x)))

summary(data[, risk_factors])

continuous_risks <- c("Albumin", "CRP", "NLR", "CallyIndex")
plot_colors <- rainbow(4)

par(mfrow = c(1,4))

for (i in 1:length(continuous_risks)) {
  var_name <- continuous_risks[i]
  var_data <- data[[var_name]]
  
  boxplot(var_data,
          main = paste("Box Plot of", var_name),
          ylab = var_name,
          col = plot_colors[i])
}

par(mfrow = c(1,1))

frequencies <- table(data$ComorbidityScore)
barplot(
    frequencies,
    main = "Frequency of Comorbidity Score (ComorbidityScore)",
    xlab = "Comorbidity Score",
    ylab = "Number of Patients",
    col = rainbow(6),
    ylim = c(0, max(frequencies) * 1.1)
    )

```



