---
title: "Data Processing"
author: "Team 2"
date: "2025-11-03"
output: pdf_document
---


```{r, eco=FALSE}

data <- read.csv("00_ProstateCancer_Data.csv", header=T)

```

```{r, eco=FALSE}

cols1 <- c("Tani_Tarihi", "Tedavi_Tarihi", "PSA_Takip_3ay", "PSA_Takip_6ay", "PSA_Takip_12ay", "BCR_Tarihi", "Metastaz_Tarihi", "Son_Takip_Tarihi")

data <- data[, !(names(data) %in% cols1)]

```

```{r, eco=FALSE}

cols2 <- c("Klinik_Evre", "Biyopsi_Gleason", "Risk_Grubu", "Komorbidite_Skor", "Tedavi_Tipi", "ADT_Tipi", "Patolojik_Evre", "Cerrahi_Sinir", "Final_Gleason", "BCR_Durum", "Metastaz_Durum", "Son_Durum")

data[cols2] <- lapply(data[cols2], as.factor)

```

```{r, eco=FALSE}

names(data) <- c("Patient ID", "Age", "PSA_before", "CTstage", "GleasonScore_before", "RiskClass", "Albumin", "Lymphocyte", "CRP", "NLR", "CallyIndex", "ComorbidityScore", "Treatment", "RadiationDose", "HormoneType", "HormonDuration" ,"TumorSize", "MarginStatus", "GleasonScore_after", "PSA_after", "BCR", "Metastasis", "Survival")

```



## Data Analysis 1: Chi-squared Test, Treatment vs BCR

### Hypotheses

H0: Treatment Type and Biochemical Recurrence (BCR) Status are independent. (The recurrence rate is the same across all treatment groups)

Ha: Treatment Type and Biochemical Recurrence (BCR) Status are not independent. (The recurrence rate is significantly different for at least one treatment group)

```{r, eco=FALSE}

bcr_table <- table(data$Treatment, data$BCR)
cat("Frequency Table (Counts):\n")
print(bcr_table)

bcr_test <- chisq.test(bcr_table)
cat("\nChi-squared Test Results:\n")
print(bcr_test)

cat("\nBCR Rate (%) within each Treatment Group:\n")
round(prop.table(bcr_table, margin = 1) * 100, 1)

```

**Conclusion**: Since the p-value (0.0294) is less than the significance level (a=0.05), we reject the null hypothesis (H0).

**Interpretation**: There is a statistically significant association between the type of treatment a patient receives and the likelihood of experiencing Biochemical Recurrence (BCR).

Highest Recurrence Rate: Treatment Type 3 (Hormone Therapy (ADT) Monotherapy) showed the highest biochemical recurrence rate at 43.7%.

Lowest Recurrence Rate: Treatment Type 1 (Radical Prostatectomy) showed the lowest recurrence rate at 25.6%.

```{r, eco=FALSE}

bcr_prop_table <- prop.table(bcr_table, margin = 1)

par(mar = c(5, 4, 4, 10) + 0.1)
barplot(
  t(bcr_prop_table) * 100,
  main = "Biochemical Recurrence (BCR) Rate by Treatment Type",
  xlab = "Treatment Type",
  ylab = "Percentage (%)",
  col = c("blue", "red"),
  legend.text = c("No Recurrence", "Recurrence"),
  args.legend = list(
    title = "Class", 
    x = "topright",
    inset = c(-0.5, 0),
    xpd = TRUE,
    cex = 0.8
  ),
  ylim = c(0, 100),
  names.arg = c("RP", "RT", "ADT", "RT + ADT")
)

```



## Data Analysis 2: Chi-squared Test, Treatment vs Survival

### Hypotheses

H0: Treatment Type and Survival Status are independent. (The survival rate is the same across all treatment groups)

Ha: Treatment Type and Survival Status are not independent. (The survival rate is significantly different for at least one treatment group)

```{r, eco=FALSE}

survival_table <- table(data$Treatment, data$Survival)
cat("Frequency Table (Counts):\n")
print(survival_table)

survival_test <- chisq.test(survival_table)
cat("\nChi-squared Test Results:\n")
print(survival_test)

cat("\nSurvival Rate (%) within each Treatment Group (Survival = Alive):\n")
round(prop.table(survival_table, margin = 1) * 100, 1)

```

**Conclusion**: Since the p-value (0.2034) is greater than the significance level (a=0.05), we fail to reject the null hypothesis (H0).

**Interpretation**: There is no statistically significant association between the type of treatment a patient receives and the likelihood of their Survival Status (Alive vs Deceased).

```{r, eco=FALSE}

survival_prop_table <- prop.table(survival_table, margin = 1)

par(mar = c(5, 4, 4, 10) + 0.1)
barplot(
  t(survival_prop_table) * 100,
  main = "Survival Rate by Treatment Type",
  xlab = "Treatment Type",
  ylab = "Percentage (%)",
  col = c("blue", "red"),
  legend.text = c("Alive", "Deceased"),
  args.legend = list(
    title = "Class", 
    x = "topright",
    inset = c(-0.5, 0),
    xpd = TRUE,
    cex = 0.8
  ),
  ylim = c(0, 100),
  names.arg = c("RP", "RT", "ADT", "RT + ADT")
)

```



## Data Analysis 3: ANOVA Test, Treatment vs PSA Difference

### Hypotheses

H0: The mean change in PSA (Delta PSA) is the same across all four Treatment Type groups

Ha: The mean change in PSA (Delta PSA) is significantly different for at least one treatment group

```{r, eco=FALSE}

data$Delta_PSA <- data$PSA_before - data$PSA_after

anova_result <- aov(Delta_PSA ~ Treatment, data = data)
summary(anova_result)

TukeyHSD(anova_result)

```

**Conclusion**: Since the p-value (2.66e-06) is less than the significance level (a=0.05), we reject the null hypothesis (H0).

**Interpretation**: Interpretation: There is a statistically significant difference in the mean change in PSA (Delta PSA) among the different treatment groups.

**Tukey's HSD Interpretation**:

The mean PSA change (Delta PSA, where a larger value indicates a greater reduction) in the Radical Prostatectomy (Treatment 1) group is significantly smaller than the mean change in all other treatment groups (2, 3, and 4).

Largest Mean Difference: Treatment 3 (Hormone Therapy Monotherapy) showed the largest positive mean difference (Delta PSA = 29.755) compared to Treatment 1, indicating the greatest average PSA reduction after treatment.

No Significant Difference: There is no significant difference in the mean PSA Difference among the three non-surgical groups (Treatment 2, 3, and 4).


```{r, eco=FALSE}

par(mar = c(5, 4, 4, 2) + 0.1)
boxplot(Delta_PSA ~ Treatment,
        data = data,
        main = "Delta PSA Distribution by Treatment Type",
        xlab = "Treatment Type",
        ylab = "Delta PSA (PSA at diagnosis - PSA after treatment)",
        col = rainbow(4),
        names = c("RP", "RT", "ADT", "RT + ADT")
)

means <- tapply(data$Delta_PSA, data$Treatment, mean, na.rm=TRUE)
points(means, pch = 8, col = "black", cex = 1.5)

```
